---
title: Exploratory data analysis II
subtitle: Lecture 6
date: 2026-09-14
format: 
  revealjs: 
    output-file: 06-exploratory-data-analysis-II-slides.html
    footer: "[🔗 sta199-f26.github.io](https://sta199-f26.github.io/)"
    theme: slides.scss
    transition: fade
    slide-number: true
    logo: "slides/images/logo.png" 
    toc: false
  html: 
    code-link: true
---

```{r}
#| label: setup
#| include: false
ggplot2::theme_set(ggplot2::theme_gray(base_size = 16))
```

# Warm up

## While you wait: Get your application exercise

If you haven't done so at the end of last class:

::: appex
-   Go to your `ae` project in Positron. 

-   Make sure all of your changes up to this point are committed and pushed, i.e., there's nothing left in your source control pane.

-   Click sync to get today's application exercise file. You can also click on the three dots next to CHANGES and click Pull.
:::

## Announcements

TO DO: Add announcements.


# From last time: Application exercise

## AE 02 {.smaller}

::: appex
Work through the application exercise `ae-02`. Periodically, and at the end, preview, commit, and push your edits.
:::

## Multivariate analysis {.smaller}

Analyzing the relationship between multiple variables:

::: incremental
-   In general, one variable is identified as the **outcome** of interest

-   The remaining variables are **predictors** or **explanatory variables**

-   Plots for exploring multivariate relationships are the same as those for bivariate relationships, but **conditional** on one or more variables

    -   Conditioning can be done via faceting or aesthetic mappings (e.g., scatterplot of `y` vs. `x1`, colored by `x2`, faceted by `x3`)

-   Summary statistics for exploring multivariate relationships are the same as those for bivariate relationships, but **conditional** on one or more variables

    -   Conditioning can be done via grouping (e.g., correlation between `y` and `x1`, grouped by levels of `x2` and `x3`)
:::

# `group_by()`, `summarize()`, `count()`

## From the application exercise: `flip_22`

```{r}
#| include: false
library(tidyverse)
gerrymander <- read_csv("data/gerrymander.csv")
gerrymander_24 <- gerrymander |> filter(!is.na(house_rep_24))
```

```{r}
gerrymander_24 <- gerrymander_24 |>
  mutate(
    flip_22_24 = case_when(
      house_party_22 == "Republican" &
        house_party_24 == "Democrat" ~ "Flip R to D",
      house_party_22 == "Democrat" &
        house_party_24 == "Republican" ~ "Flip D to R",
      .default = "No Flip"
    ),
    gerry_26_cat = case_when(
      gerry_26 %in% c("A", "B") ~ "Low",
      gerry_26 == "C" ~ "Medium",
      gerry_26 %in% c("D", "F") ~ "High"
    )
  ) |>
  relocate(house_party_22, house_party_24, flip_22_24, gerry_26, .after = state)
```

## Spot the difference {.smaller}

::: question
What does `group_by()` do?
:::

::: columns
::: column

```{r}
gerrymander_24 |>
  filter(!is.na(gerry_26_cat)) |>
  count(flip_22_24, gerry_26_cat)
```

:::
::: column

```{r}
gerrymander_24 |>
  filter(!is.na(gerry_26_cat)) |>
  count(flip_22_24, gerry_26_cat) |>
  group_by(flip_22_24)
```

:::
:::

## Let's simplify! {.smaller}

::: question
What does `group_by()` do in the following pipeline?
:::

::: columns
::: column

```{r}
gerrymander |>
  summarize(mean_trump_24 = mean(trump_24, na.rm = TRUE))
```

:::
::: column

```{r}
gerrymander |>
  group_by(state) |>
  summarize(mean_trump_24 = mean(trump_24, na.rm = TRUE))
```

:::
:::

## `group_by()` {.smaller}

-   Group by converts a data frame to a grouped data frame, where subsequent operations are performed once per group

-   `ungroup()` removes grouping

::: columns
::: column

```{r}
gerrymander_24 |>
  select(state, district, house_party_22, house_party_24) |>
  group_by(state)
```

:::
::: {.column .fragment}

```{r}
#| code-line-numbers: "4"
gerrymander_24 |>
  select(state, district, house_party_22, house_party_24) |>
  group_by(state) |>
  ungroup()
```

:::
:::

## `group_by() |> summarize()` {.smaller}

A common pipeline is `group_by()` and then `summarize()` to **calculate** summary statistics for each group:

```{r}
gerrymander_24 |>
  group_by(state) |>
  summarize(
    mean_trump_24 = mean(trump_24, na.rm = TRUE),
    median_trump_24 = median(trump_24, na.rm = TRUE)
  )
```

## `group_by() |> summarize()` {.smaller}

This pipeline can also be used to **count** number of observations for each group:

```{r}
gerrymander_24 |>
  group_by(state) |>
  summarize(n = n())
```

## `summarize()` {.smaller}

```{r}
#| eval: false
... |>
  summarize(
    name_of_summary_statistic = summary_function(variable)
  )
```

. . .

-   `name_of_summary_statistic`: Anything you want to call it!
    -   Recommendation: Keep it short and evocative
-   `summary_function()`:
    -   `n()`: number of observations
    -   `mean()`: mean
    -   `median()`: median
    -   `sd()`: standard deviation
    -   `min()`: minimum
    -   `max()`: maximum
    -   `IQR()`: inter-quartile range
    -   `quantile()`: quantile (e.g., 0.25, 0.75, etc.)

## Spot the difference {.smaller}

::: question
What's the difference between the following two pipelines?
:::

::: columns

::: column

```{r}
gerrymander_24 |>
  group_by(state) |>
  summarize(n = n())
```

:::

::: column

```{r}
gerrymander_24 |>
  count(state)
```

:::

:::

## `count()`

::: columns
::: {.column width="40%"}
``` r
... |>
  count(variable)
```
:::

::: {.column width="60%"}

```{r}
#| eval: false
... |>
  count(variable1, variable2)
```

:::
:::

-   Count the number of observations in each level of variable(s)

-   Place the counts in a variable called `n`

## Review {.xsmall}

::: task
How would you write the following pipeline with `count()` instead?

```{r}
#| output-location: column
gerrymander_24 |>
  group_by(state) |>
  summarize(n = n()) |>
  arrange(desc(n))
```

::: wooclap-choices
- `gerrymander_24 |> arrange(state) |> count()`
- `gerrymander_24 |> count(state) |> arrange(desc(n))`
- `gerrymander_24 |> count(state) |> sort(n)`
- `gerrymander_24 |> count(state, sort = TRUE)`
:::

:::

# `mutate()`

## Flip the question

::: question
How does the prevalence of gerrymandering for the 2026 election vary among Congressional Districts with different flip outcomes between the 2022 and 2024 elections?
:::

vs.

::: question
How does the distribution of 2022–2024 flip outcomes vary among Congressional Districts with low, medium, or high prevalence of gerrymandering for the 2026 election?
:::

## What's going on? {.smaller}

::: {.question .xsmall}
The following code plots the distribution of 2022–2024 flip outcomes within each level of gerrymandering for the 2026 election. However, the gerrymandering levels appear in an unexpected order.
What's going on?
:::

```{r}
gerrymander_24 |>
  filter(!is.na(gerry_26_cat)) |>
  ggplot(aes(x = gerry_26_cat, fill = flip_22_24)) +
  geom_bar(position = "fill")
```

## Another glimpse at `gerrymander_24` {.smaller}

```{r}
gerrymander_24 |>
  select(
    district,
    house_party_22,
    house_party_24,
    flip_22_24,
    gerry_26,
    gerry_26_cat
  ) |>
  glimpse()
```

## `mutate()`

-   We want to display `gerry_26_cat` in the order Low, Medium, High

-   But it's stored as a character variable, so the plot orders its values alphabetically

-   We can convert it to a factor with explicitly specified levels to control the order

-   The `mutate()` function transforms (*mutates*) a data frame by creating a new column or updating an existing one

## `mutate()` in action {.smaller}

You can create a new variable with `mutate()`:

```{r}
#| code-line-numbers: "|2-6|7"
gerrymander_24 |>
  mutate(
    gerry_26_factor = fct_relevel(gerry_26_cat, "Low", "Medium", "High")
  ) |>
  relocate(district, gerry_26_cat, gerry_26_factor) # relocate for easier viewing
```

## Revisit the plot {.smaller}

::: question
How does the distribution of 2022–2024 flip outcomes vary by prevalence of gerrymandering for the 2026 election?
:::

```{r}
#| output-location: column
#| code-line-numbers: "|2-6|8"
gerrymander_24 |>
  mutate(
    gerry_26_factor = factor(
      gerry_26_cat,
      levels = c("Low", "Medium", "High")
    )
  ) |>
  filter(!is.na(gerry_26_factor)) |>
  ggplot(aes(x = gerry_26_factor, fill = flip_22_24)) +
  geom_bar(position = "fill")
```

## `mutate()` and overwrite

You can overwrite an existing variable with `mutate()`:

```{r}
gerrymander_24 |>
  mutate(
    gerry_26_cat = factor(gerry_26_cat, levels = c("Low", "Medium", "High"))
  ) |>
  select(district, gerry_26_cat)
```

## `mutate()` and `if_else()` {.smaller}

Use `mutate()` with `if_else()` to recode with an either/or logic:

> If `flip_22_24` is "Flip R to D", recode it as "Democrat flipped", otherwise recode it as "Other outcome".

```{r}
#| code-line-numbers: "|2-6"
gerrymander_24 |>
  mutate(
    dem_flip_22_24 = if_else(
      flip_22_24 == "Flip R to D",
      "Democrat flipped",
      "Other outcome"
    )
  ) |>
  select(district, flip_22_24, dem_flip_22_24)
```

## `mutate()` and `case_when()` {.smaller}

Use `mutate()` with `case_when()` to recode with a more complex logic:

> For the 2026 election, recode gerrymandering grades A or B as "Low", C as "Medium", and D or F as "High".

```{r}
gerrymander_24 |>
  mutate(
    gerry_26_cat = case_when(
      gerry_26 %in% c("A", "B") ~ "Low",
      gerry_26 == "C" ~ "Medium",
      gerry_26 %in% c("D", "F") ~ "High"
    )
  ) |>
  select(district, gerry_26, gerry_26_cat) |>
  group_by(gerry_26) |> # group by grade
  slice_head(n = 1) # show top row per group
```

Grades that don't match any condition, including missing grades, get `NA`.

## `mutate()` and store {.smaller}

If you want to keep your changes, you need to store the data frame after `mutate()`:

::: {.columns}
::: {.column}

```{r}
#| error: true
gerrymander_24 |>
  mutate(dem_flip_22_24 = flip_22_24 == "Flip R to D") |>
  select(district, dem_flip_22_24) |>
  slice_head(n = 3) # show top 3 rows

gerrymander_24 |>
  select(district, dem_flip_22_24)
```

:::
::: {.column .fragment}

```{r}
gerrymander_24 <- gerrymander_24 |>
  mutate(dem_flip_22_24 = flip_22_24 == "Flip R to D")

gerrymander_24 |>
  select(district, dem_flip_22_24) |>
  slice_head(n = 3) # show top 3 rows
```

:::
:::
