---
title: Exploratory data analysis II
subtitle: Lecture 6
date: 2026-09-14
format: 
  revealjs: 
    output-file: 06-exploratory-data-analysis-II-slides.html
    footer: "[🔗 sta199-f26.github.io](https://sta199-f26.github.io/)"
    theme: slides.scss
    transition: fade
    slide-number: true
    logo: "slides/images/logo.png" 
    toc: false
  html: 
    code-link: true
---

```{r}
#| label: setup
#| include: false
ggplot2::theme_set(ggplot2::theme_gray(base_size = 16))
```

# Warm up

## While you wait: Get your application exercise

If you haven't done so at the end of last class:

::: appex
-   Go to your `ae` project in Positron. 

-   Make sure all of your changes up to this point are committed and pushed, i.e., there's nothing left in your source control pane.

-   Click sync to get today's application exercise file. You can also click on the three dots next to CHANGES and click Pull.
:::

## Announcements

TO DO: Add announcements.


# From last time: Application exercise

## AE 02 {.smaller}

::: appex
Work through the application exercise `ae-02`. Periodically, and at the end, preview, commit, and push your edits.
:::

## Multivariate analysis {.smaller}

Analyzing the relationship between multiple variables:

::: incremental
-   In general, one variable is identified as the **outcome** of interest

-   The remaining variables are **predictors** or **explanatory variables**

-   Plots for exploring multivariate relationships are the same as those for bivariate relationships, but **conditional** on one or more variables

    -   Conditioning can be done via faceting or aesthetic mappings (e.g., scatterplot of `y` vs. `x1`, colored by `x2`, faceted by `x3`)

-   Summary statistics for exploring multivariate relationships are the same as those for bivariate relationships, but **conditional** on one or more variables

    -   Conditioning can be done via grouping (e.g., correlation between `y` and `x1`, grouped by levels of `x2` and `x3`)
:::


## From the application exercise: `flip_22_24`

```{r}
#| include: false
library(tidyverse)
gerrymander <- read_csv("data/gerrymander.csv")
gerrymander_24 <- gerrymander |> filter(!is.na(house_rep_24))
```

```{r}
gerrymander_24 <- gerrymander_24 |>
  mutate(
    flip_22_24 = case_when(
      house_party_22 == "Republican" &
        house_party_24 == "Democrat" ~ "Flip R to D",
      house_party_22 == "Democrat" &
        house_party_24 == "Republican" ~ "Flip D to R",
      .default = "No Flip"
    )
  ) |>
  relocate(house_party_22, house_party_24, flip_22_24, gerry_26, .after = state)
```

# `mutate()`

## From grades to categories

::: question
How can we combine the 2026 gerrymandering grades into low, medium, and high prevalence of gerrymandering?
:::

-   A or B → Low
-   C → Medium
-   D or F → High

We'll create a new variable called `gerry_26_cat` from `gerry_26`.

## `mutate()`

-   The `mutate()` function transforms (*mutates*) a data frame by creating a new column or updating an existing one

-   We choose a name for the column on the left of `=` and describe how to calculate it on the right

-   To create `gerry_26_cat`, we need different values for different grades

```{r}
#| eval: false
gerrymander_24 |>
  mutate(gerry_26_cat = ...)
```

## `mutate()` and `case_when()` {.smaller}

Use `case_when()` to recode with multiple conditions:

```{r}
#| code-line-numbers: "|2-8|4-6|9"
gerrymander_24 |>
  mutate(
    gerry_26_cat = case_when(
      gerry_26 %in% c("A", "B") ~ "Low",
      gerry_26 == "C" ~ "Medium",
      gerry_26 %in% c("D", "F") ~ "High"
    )
  ) |>
  select(district, gerry_26, gerry_26_cat)
```

-   Each condition goes on the left of `~`, and its new value goes on the right
-   The first matching condition determines the value
-   Grades that don't match any condition, including missing grades, get `NA`

## `mutate()` and store {.smaller}

Creating a column in a pipeline doesn't change the original data frame:

```{r}
#| error: true
gerrymander_24 |>
  select(district, gerry_26, gerry_26_cat)
```

. . .

To keep the new column, store the result:

```{r}
gerrymander_24 <- gerrymander_24 |>
  mutate(
    gerry_26_cat = case_when(
      gerry_26 %in% c("A", "B") ~ "Low",
      gerry_26 == "C" ~ "Medium",
      gerry_26 %in% c("D", "F") ~ "High"
    )
  )
```

## Check the new variable {.smaller}

```{r}
gerrymander_24 |>
  select(district, gerry_26, gerry_26_cat)
```

::: question
What value does `gerry_26_cat` have when `gerry_26` is missing? Why?
:::

## What's going on? {.smaller}

::: question
The gerrymandering categories appear in an unexpected order. What's going on?
:::

```{r}
gerrymander_24 |>
  filter(!is.na(gerry_26_cat)) |>
  ggplot(aes(x = flip_22_24, fill = gerry_26_cat)) +
  geom_bar(position = "fill")
```

## `mutate()` and `fct_relevel()` {.smaller}

-   `flip_22_24` is a character variable, so the plot orders its values alphabetically
-   `fct_relevel()` creates a factor with levels in the specified order
-   We can use `mutate()` to overwrite `flip_22_24` and store the change

```{r}
#| code-line-numbers: "|1|3"
gerrymander_24 <- gerrymander_24 |>
  mutate(
    gerry_26_cat = fct_relevel(gerry_26_cat, "Low", "Medium", "High")
  )

gerrymander_24 |>
  select(district, gerry_26, gerry_26_cat)
```

## Revisit the plot {.smaller}

::: question
How does the distribution of 2022–2024 flip outcomes vary by prevalence of gerrymandering for the 2026 election?
:::

```{r}
gerrymander_24 |>
  filter(!is.na(gerry_26_cat)) |>
  ggplot(aes(x = flip_22_24, fill = gerry_26_cat)) +
  geom_bar(position = "fill")
```

# `group_by()`, `summarize()`, `count()`

## Spot the difference {.smaller}

::: question
What does `group_by()` do?
:::

::: columns
::: column

```{r}
gerrymander_24 |>
  filter(!is.na(gerry_26_cat)) |>
  count(flip_22_24, gerry_26_cat)
```

:::
::: column

```{r}
gerrymander_24 |>
  filter(!is.na(gerry_26_cat)) |>
  count(flip_22_24, gerry_26_cat) |>
  group_by(flip_22_24)
```

:::
:::

## Let's simplify! {.smaller}

::: question
What does `group_by()` do in the following pipeline?
:::

::: columns
::: column

```{r}
gerrymander |>
  summarize(mean_trump_24 = mean(trump_24, na.rm = TRUE))
```

:::
::: column

```{r}
gerrymander |>
  group_by(state) |>
  summarize(mean_trump_24 = mean(trump_24, na.rm = TRUE))
```

:::
:::

## `group_by()` {.smaller}

-   Group by converts a data frame to a grouped data frame, where subsequent operations are performed once per group

-   `ungroup()` removes grouping

::: columns
::: column

```{r}
gerrymander_24 |>
  select(state, district, house_party_22, house_party_24) |>
  group_by(state)
```

:::
::: {.column .fragment}

```{r}
#| code-line-numbers: "4"
gerrymander_24 |>
  select(state, district, house_party_22, house_party_24) |>
  group_by(state) |>
  ungroup()
```

:::
:::

## `group_by() |> summarize()` {.smaller}

A common pipeline is `group_by()` and then `summarize()` to **calculate** summary statistics for each group:

```{r}
gerrymander_24 |>
  group_by(state) |>
  summarize(
    mean_trump_24 = mean(trump_24, na.rm = TRUE),
    median_trump_24 = median(trump_24, na.rm = TRUE)
  )
```

## `group_by() |> summarize()` {.smaller}

This pipeline can also be used to **count** number of observations for each group:

```{r}
gerrymander_24 |>
  group_by(state) |>
  summarize(n = n())
```

## `summarize()` {.smaller}

```{r}
#| eval: false
... |>
  summarize(
    name_of_summary_statistic = summary_function(variable)
  )
```

. . .

-   `name_of_summary_statistic`: Anything you want to call it!
    -   Recommendation: Keep it short and evocative
-   `summary_function()`:
    -   `n()`: number of observations
    -   `mean()`: mean
    -   `median()`: median
    -   `sd()`: standard deviation
    -   `min()`: minimum
    -   `max()`: maximum
    -   `IQR()`: inter-quartile range
    -   `quantile()`: quantile (e.g., 0.25, 0.75, etc.)

## Spot the difference {.smaller}

::: question
What's the difference between the following two pipelines?
:::

::: columns

::: column

```{r}
gerrymander_24 |>
  group_by(state) |>
  summarize(n = n())
```

:::

::: column

```{r}
gerrymander_24 |>
  count(state)
```

:::

:::

## `count()`

::: columns
::: {.column width="40%"}
``` r
... |>
  count(variable)
```
:::

::: {.column width="60%"}

```{r}
#| eval: false
... |>
  count(variable1, variable2)
```

:::
:::

-   Count the number of observations in each level of variable(s)

-   Place the counts in a variable called `n`

## Review {.xsmall}

::: task
How would you write the following pipeline with `count()` instead?

```{r}
#| output-location: column
gerrymander_24 |>
  group_by(state) |>
  summarize(n = n()) |>
  arrange(desc(n))
```

::: wooclap-choices
- `gerrymander_24 |> arrange(state) |> count()`
- `gerrymander_24 |> count(state) |> arrange(desc(n))`
- `gerrymander_24 |> count(state) |> sort(n)`
- `gerrymander_24 |> count(state, sort = TRUE)`
:::

:::

