---
title: Exam 1 Review
subtitle: Lecture 12
author: Prof. Mine Çetinkaya-Rundel
date: 2026-09-30
format: 
  revealjs: 
    output-file: 12-exam-I-review-slides.html
    footer: "[🔗 sta199-f26.github.io](https://sta199-f26.github.io/)"
    theme: slides.scss
    transition: fade
    slide-number: true
    logo: "/slides/images/logo.png"
    toc: false
  html: 
    code-link: true
filters:
  - ../remove-fmt-skip.lua
---

```{r}
#| label: setup
#| include: false
ggplot2::theme_set(ggplot2::theme_gray(base_size = 16))
```

```{r}
#| include: false
library(tidyverse)
```

```{r}
#| label: review-data
#| include: false

# Match the preparation in exam-review/exam-1-review.qmd.
blizzard_salary <- openintro::blizzard_salary |>
  mutate(
    annual_salary = case_when(
      salary_type == "week" ~ current_salary * 52,
      salary_type == "hour" ~ current_salary * 40 * 52,
      TRUE ~ current_salary
    ),
    performance_rating = if_else(
      performance_rating == "Developing",
      "Poor",
      performance_rating
    )
  ) |>
  filter(salary_type != "week") |>
  mutate(salary_type = if_else(salary_type == "hour", "Hourly", "Salaried")) |>
  filter(!is.na(annual_salary)) |>
  select(percent_incr, salary_type, annual_salary, performance_rating)

penguins <- palmerpenguins::penguins
flights <- nycflights13::flights

tv <- read_csv(here::here("exam-review", "data", "tv.csv"))
romance_shows <- tv |>
  filter(genre_1 == "Romance" | genre_2 == "Romance" | genre_3 == "Romance") |>
  distinct(title) |>
  pull(title)
comedy_shows <- tv |>
  filter(genre_1 == "Comedy" | genre_2 == "Comedy" | genre_3 == "Comedy") |>
  distinct(title) |>
  pull(title)

# Preserve the review's plotted labels so Q25 matches the question and key.
# The source reverses the genre labels; see the instructor note on Q25.
romance_comedy_means <- tv |>
  filter(title %in% c(romance_shows, comedy_shows)) |>
  filter(!(title %in% romance_shows & title %in% comedy_shows)) |>
  mutate(genre = if_else(title %in% romance_shows, "Comedy", "Romance")) |>
  group_by(genre, title) |>
  summarize(mean_av_rating = mean(av_rating), .groups = "drop")

movies <- read_csv(here::here("exam-review", "data", "movies.csv")) |>
  mutate(
    rating = case_when(
      rating == "TV-PG" ~ "PG",
      rating == "Unrated" ~ "Not Rated",
      is.na(rating) ~ "Not Rated",
      .default = rating
    ),
    rating = fct_relevel(rating, "G", "PG", "PG-13", "R", "NC-17", "Not Rated"),
    genre = fct_other(genre, keep = c("Action", "Comedy", "Drama", "Horror")),
    genre = fct_relevel(genre, "Other", "Horror", "Drama", "Action", "Comedy")
  )

country_inflation <- read_csv(here::here(
  "exam-review",
  "data",
  "country-inflation.csv"
))
country_inflation_long <- country_inflation |>
  pivot_longer(
    cols = !country,
    names_to = "year",
    names_transform = as.numeric,
    values_to = "inflation_rate"
  )
inflation_ratios <- country_inflation |>
  filter(!is.na(`1993`), !is.na(`2025`)) |>
  mutate(inf_ratio = `2025` / `1993`) |>
  select(country, inf_ratio) |>
  arrange(desc(inf_ratio))
```

# Warm-up

## While you wait: Participate 📱💻 {.xsmall}

- Check your email (and your spam folder) for an email from TEAMMATES

- Fill out your peer evaluation for your team members

::: caution-warning
If you fail to submit your feedback, you're not eligible to receive the points they give you.
:::

## From last time {.smaller}

::: appex
-   Go to your ae project in Positron.

-   If you haven't yet done so, make sure all of your changes up to this point are committed and pushed, i.e., there's nothing left in your source control pane.

-   If you haven't yet done so, pull to get today's application exercise file: `ae-07-chronicle-scrape-I.qmd`.

-   Work through the application exercise in class, and render, commit, and push your edits by the end of class.
:::

# Exam 1 review

## Exam 1 {.smaller}

Exam 1 is on Wednesday, starts promptly at 1:25 pm!

- Covers all material up to and including today's lecture

- Sample exam + model solutions for select HW questions are posted

- Exam will be in class, closed book, no calculators, no computers, just 1 notes sheet

- Notes sheet: 8.5x11, both sides, hand written or typed, any content you want, must be prepared by you

- We'll direct you to your seat so get here early to allow time to get settled and read the instructions

## Types of questions {.smaller}

- Multiple-choice:
  - Some will be single best answer, no partial credit
  - Some will be *at least one* correct answer, with partial credit, and clearly identified with "Select all that apply"

- Open-ended: 
  - Short answer, show your work, justify your reasoning
  - With partial credit

## Q1: Variable types {.xsmall}

`blizzard_salary` has **409 rows and 4 columns**. Each row is a worker who completed the spreadsheet. Which statements are correct? **Select all that apply.**

a. The dataset has 399 rows.
b. The dataset has 4 columns. [**Correct**]{.fragment}
c. Each row represents a Blizzard Entertainment worker who filled out the spreadsheet. [**Correct**]{.fragment}
d. `percent_incr` is numerical and discrete.
e. `salary_type` is numerical.
f. `annual_salary` is numerical. [**Correct**]{.fragment}
g. `performance_rating` is categorical and ordinal. [**Correct**]{.fragment}

## Q9: Where did the observations go? — question {.xsmall}

A friend notices that the total number of observations in this plot seems lower than the total in `blizzard_salary` (**409 workers**). What might be going on here? Explain your reasoning.

```{r}
#| label: q9-question-plot
#| echo: false
#| fig-width: 6
#| fig-height: 3.8
#| out-width: 100%
#| fig-alt: Stacked counts of workers by salary type and performance rating; both bars together contain fewer than 409 workers.
salary_rating_plot <- blizzard_salary |>
  drop_na(salary_type, performance_rating) |>
  ggplot(aes(x = salary_type, fill = fct_rev(performance_rating))) +
  geom_bar() +
  scale_fill_viridis_d() +
  labs(x = "Salary type", y = "Count", fill = "Performance\nrating")
salary_rating_plot
```

[**Observations with missing values for both variables won't be represented on the plot.**]{.fragment}

## Q13: Reviewing an interpretation — question {.xsmall}



::: columns

::: {.column width="75%"}

For **Top performers**, a team writes: “The relationship is positive, having a higher salary results in a higher percent increase. There is one clear outlier.” Which peer review notes are accurate and helpful? **Select all that apply.**

:::

::: {.column width="25%"}

```{r}
#| label: q13-top-plot
#| echo: false
#| fig-width: 5.2
#| fig-height: 4.2
#| out-width: 100%
#| fig-alt: Salary versus percent increase for Top performers, showing a positive roughly linear association with substantial scatter.
top_performers_plot <- blizzard_salary |>
  filter(performance_rating == "Top") |>
  ggplot(aes(x = annual_salary, y = percent_incr)) +
  geom_point(na.rm = TRUE) +
  scale_x_continuous(
    labels = scales::label_dollar(scale = 1 / 1000, suffix = "K")
  ) +
  labs(x = "Annual salary", y = "Percent increase", title = "Top performers")
top_performers_plot
```

:::

:::

a. The interpretation is complete and perfect.
b. It doesn't mention the direction.
c. It doesn't mention the form, which is linear. [**Correct.**]{.fragment}
d. It doesn't mention the strength, which is somewhat strong. [**Correct.**]{.fragment}
e. There isn't a clear outlier; identify any potential outlier by its salary and/or percent increase. [**Correct.**]{.fragment}
f. The interpretation is causal; observational data do not establish which variable causes the other or rule out other explanations. [**Correct.**]{.fragment}

## Q16: Which bar plot? {.xsmall}

Which plot is produced by this code? [**Correct: D.**]{.fragment}

```{r}
#| eval: false
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar()
```

```{r}
#| label: q16-options
#| echo: false
#| fig-width: 12
#| fig-height: 5.4
#| out-width: 95%
#| fig-alt: "Five options: A island proportions filled by species; B species proportions filled by island; C horizontal species proportions; D island counts filled by species; E horizontal species counts filled by island."
q16_options <- list(
  ggplot(penguins, aes(x = island, fill = species)) +
    geom_bar(position = "fill", color = "black") +
    labs(title = "A", x = "Island", y = "Proportion", fill = "Species"),
  ggplot(penguins, aes(x = species, fill = island)) +
    geom_bar(position = "fill", color = "black") +
    labs(title = "B", x = "Species", y = "Proportion", fill = "Island"),
  ggplot(penguins, aes(y = species, fill = island)) +
    geom_bar(position = "fill", color = "black") +
    labs(title = "C", y = "Species", x = "Proportion", fill = "Island"),
  ggplot(penguins, aes(x = island, fill = species)) +
    geom_bar(color = "black") +
    labs(title = "D", x = "Island", y = "Count", fill = "Species"),
  ggplot(penguins, aes(y = species, fill = island)) +
    geom_bar(color = "black") +
    labs(title = "E", y = "Species", x = "Count", fill = "Island")
)
q16_options <- map(
  q16_options,
  ~ .x +
    scale_fill_grey(start = 0.8, end = 0) +
    theme_gray(base_size = 12) +
    theme(
      legend.position = "bottom",
      legend.title = element_blank(),
      legend.text = element_text(size = 8),
      legend.key.size = unit(0.25, "cm")
    )
)
patchwork::wrap_plots(q16_options, ncol = 3)
```

## Q17: Reading a tibble {.xsmall}

Based on this output, which statements **must be true**? **Select all that apply.**

::: columns
::: {.column width="60%"}

```{r}
#| label: q17-flights-peek
#| echo: false
print(
  flights |> relocate(year, month, day, arr_delay, carrier),
  n = 10,
  width = 65
)
```

:::
::: {.column width="40%"}

a. `flights` is a tibble. [**Correct.**]{.fragment}
b. `flights` has 10 rows.
c. `flights` has 8 columns.
d. `carrier` is a character variable. [**Correct.**]{.fragment}
e. There are no missing data in `flights`.

:::
:::

## Q27: Summing a condition {.xsmall}

Which statements about the code and its result are true? **Select all that apply.**

```{r}
#| label: q27-original-summary
movies |>
  summarize(sum(release_country == "United States"))
```

a. It evaluates whether each `release_country` equals `"United States"`, resulting in logical values. [**Correct.**]{.fragment}
b. It filters out rows where `release_country` is not `"United States"` and counts the remaining rows.
c. It sums the logical values, treating each `TRUE` as 1 and each `FALSE` as 0. [**Correct.**]{.fragment}
d. It results in a character vector.
e. The result shows 435 movies released in the United States. [**Correct.**]{.fragment}

## Q29: A 6 × 6 result {.xsmall}

```{r}
#| label: q29-wide-result
movies |>
  count(rating, genre) |>
  pivot_wider(names_from = genre, values_from = n, values_fill = 0)
```

Which statements are true? **Select all that apply.**

a. The code counts movies in each rating and genre combination. [**Correct.**]{.fragment}
b. The code sorts the results in descending order.
c. Each row of the output is a movie.
d. The output shows six distinct ratings in the dataset. [**Correct.**]{.fragment}
e. The code reduces the number of variables and observations in the `movies` data frame to six.

## Q31: Reshaping inflation data — question {.xsmall}

`country_inflation` has **38 rows and 34 columns**: one row per country, a `country` column, and annual inflation columns for **1993–2025**.

```{r}
#| eval: false
country_inflation_long <- country_inflation |>
  pivot_longer(
    cols = !country,
    names_to = "year",
    names_transform = as.numeric,
    values_to = "inflation_rate"
  )
```

::: question
How many rows and columns does the result have? **Show where each number comes from.** What does each row represent, what did `pivot_longer()` do, and what does `names_transform = as.numeric` accomplish?
:::

## Q31: Reshaping inflation data — answer {.xsmall}

**Rows: 1,254. Columns: 3.** Each row represents **one country in one year**.

1. **38** is the number of countries (the original rows).
2. **34 − 1 = 33** columns are year columns: exclude the one `country` column. Equivalently, **2025 − 1993 + 1 = 33** years.
3. Each country becomes **33 rows**, so the total is:

$$38 \times 33 = 38 \times 30 + 38 \times 3 = 1{,}140 + 114 = 1{,}254$$

The output columns are **`country` + `year` + `inflation_rate` = 3**:

- `country` is retained and repeated for each year.
- The 33 old column **names** become values in `year`.
- Their cell **values** become values in `inflation_rate`.

Missing inflation values still get rows; this code does not drop them.

## Q32: Ratios and the top five — question {.xsmall}

```{r}
#| label: q32-original-pipeline
country_inflation |>
  filter(!is.na(`1993`), !is.na(`2025`)) |>
  mutate(inf_ratio = `2025` / `1993`) |>
  select(country, inf_ratio) |>
  arrange(desc(inf_ratio)) |>
  slice_head(n = 5)
```

**Part 1:** What does the resulting data frame represent? What does New Zealand's `inf_ratio` of 2.20 mean?

**Part 2:** What would change if the `filter()` step were removed? Why could that be misleading?

## Q32: Ratios and the top five — answer {.xsmall}

**Part 1:** These are the **five countries with the largest 2025-to-1993 inflation ratios**, in descending order.

- A ratio of **2.20** means New Zealand's 2025 inflation rate was **2.20 times its 1993 rate** (120% higher). It does not mean the 2025 inflation rate was 2.20%.
- `arrange(desc(inf_ratio))` puts the largest ratios first. **`slice_head(n = 5)` keeps the first five rows in that order.**

**Part 2:** Without `filter()`, six countries would have `NA` ratios because at least one of their two rates is missing.

- `arrange()` puts these `NA`s last. **For these data, the same five countries would still be returned.**
- The top-five table cannot reveal whether an absent country has a smaller ratio or lacks data. Filtering makes the exclusion explicit; report that the ranking uses **32 of 38 countries**.

## Q32: What exactly does `slice_head()` do? {.xsmall}

**It selects rows by their current positions. It does not sort or calculate a summary.**

| Stage | Rows | What happened? |
|:--|--:|:--|
| Start | 38 | One row per country |
| `filter(...)` | 32 | Keep countries with both endpoint rates |
| `mutate(...)` | 32 | Add one ratio to each row |
| `select(...)` | 32 | Keep two columns |
| `arrange(desc(inf_ratio))` | 32 | Put largest ratios first |
| `slice_head(n = 5)` | 5 | Keep rows 1–5 |

Removing **only `slice_head(n = 5)`** keeps all **32 eligible countries**. It does not restore the six countries excluded by `filter()`.

## Other questions {.smaller}

- [Q2](/exam-review/exam-1-review.qmd#question-2)
- [Q4](/exam-review/exam-1-review.qmd#question-4)
- [Q7](/exam-review/exam-1-review.qmd#questions-6-and-7)
- [Q8](/exam-review/exam-1-review.qmd#questions-8---10)
- [Q14](/exam-review/exam-1-review.qmd#question-14)
- [Q23](/exam-review/exam-1-review.qmd#question-23)
- [Q24](/exam-review/exam-1-review.qmd#question-24)
- [Q28](/exam-review/exam-1-review.qmd#question-28)
- [Q30](/exam-review/exam-1-review.qmd#question-30)

[Exam review answers](/exam-review/exam-1-review-A.qmd)
