---
title: HW 1
subtitle: Exploring Spotify charts
description: "Due: Wed, Sep 9, 11:59 pm"
filters:
  - ../remove-fmt-skip.lua
categories: HW
---

# Introduction

## Learning objectives

- Gain practice with
  - the data science workflow using R, Positron, Git, and GitHub,
  - writing a reproducible report using Quarto, and
  - version control using Git and GitHub
- Create data visualizations with the `tidyverse`, specifically using `ggplot2`
- Wrangle data with the `tidyverse`, specifically using `dplyr`
- Interpret data visualizations

## Getting Started

To get started, follow the instructions below.

{{< include _getting-started.qmd >}}

Open the `hw-1.qmd` template Quarto file and update the author field with your first and last name. Click `Preview` to render the document, then check the resulting PDF to confirm that your name appears correctly. 

::: source-control
Stage and commit your changes with a meaningful commit message, then select `Sync Changes` to push your updates to GitHub.
:::

## Guidelines

The guidelines should feel familiar, as they are the same ones from the lab! They are included below as a reminder.

{{< include _guidelines.qmd >}}

## AI support and feedback

{{< include _ai-support-feedback.qmd >}}

::: callout-important
Since this is your first time using Codex in Positron, you will first need to install the relevant extension and log in.
To do so,

- Click on the Extensions icon in your activity bar.
- Search for `Codex` and install the `Codex - OpenAI's coding agent` extension.

Then, log in with your Duke credentials. To do so, in the Codex window in Positron, click on "More options" and then "Use device code". This will ask if you want to open an external website, click Open, and then paste the device code provided.
:::

## Packages

You will use the **tidyverse** package for data wrangling and visualization, **scales** for better axis labels, and **ggthemes** for additional color palettes.

```{r}
#| eval: true
#| message: false
library(tidyverse)
```

# Questions

For this homework, you will use the Spotify data from [Lab 1](../lab/lab-1.qmd).

You will analyze a fixed snapshot of the first 200 songs on Spotify's Global Weekly chart for the week dated August 27, 2026. Each row represents a song.
Spotify releases a new chart every week, and a website called [Kworb.net](https://kworb.net/) republishes a snapshot of these data weekly. This week's data was retrieved from [Kworb's republished Spotify Global Weekly chart](https://kworb.net/spotify/country/global_weekly.html) on August 29, 2026.
Spotify states that its chart stream counts are generated using a formula that filters for chart eligibility, so `streams` should not be interpreted as every raw Spotify play.[^1] The data also include a broad `genre` classification. These labels describe the credited artist's general musical style, not necessarily the genre of every individual song; artists and songs can belong to more than one genre.

[^1]: See Spotify's chart documentation: <https://support.spotify.com/ee-en/artists/article/understanding-spotify-charts/>.

Read the data into R with:
```{r}
#| label: load-data
spotify <- read_csv("data/spotify-global-weekly-2026-08-27.csv")
```

## Question 1

Visualize the distribution of `weeks_on_chart` using a histogram with `geom_histogram()` with three different binwidths: a binwidth of 5, a binwidth of 50, and a binwidth of 500. For example, you can create the first plot with:

```{r}
#| eval: false
ggplot(spotify, aes(x = weeks_on_chart)) +
  geom_histogram(binwidth = 5) +
  labs(
    x = "Weeks on Chart",
    y = "Count",
    title = "Distribution of Weeks on Spotify's Global Weekly Chart",
    subtitle = "Binwidth = 5 weeks"
  )
```

You will need to make three different histograms. Make sure to set informative titles and axis labels for each of your plots. Then, comment on which binwidth is most appropriate for these data and why.

::: source-control
Preview, commit, and push your changes to GitHub with the commit message "Added answer for Question 1".

Make sure to commit and push all changed files so that your source control pane is empty afterward.
:::

## Question 2

Visualize the distribution of `weeks_on_chart` again, this time using a boxplot with `geom_boxplot()`. Make sure to set informative titles and axis labels for your plot. Then, using information as needed from the box plot as well as the histogram from Question 1, describe the distribution of `weeks_on_chart` and comment on any potential outliers. Then, use a data wrangling pipeline to identify the outliers based on a visual inspection of the plot, and then display only the `artist`, `song`, `weeks_on_chart`, and `rank` variables in your output. Display the results in descending order of `weeks_on_chart`. Finally, comment on whether any of these outliers are surprising or expected for you, and why.

::: callout-tip
In describing a distribution, make sure to mention shape, center, spread, and any unusual observations.
:::

::: source-control
Preview, commit, and push your changes to GitHub with the commit message "Added answer for Question 2".

Make sure to commit and push all changed files so that your source control pane is empty afterward.
:::

## Question 3

A song's `peak_rank` is its best-ever position during its current run on the chart. Because smaller ranks are better, create a variable called `positions_below_peak` that measures how many positions a song's current rank is below its peak rank.

```{r}
#| eval: false
spotify <- spotify |>
  mutate(positions_below_peak = rank - peak_rank)
```

In one pipeline, display the 10 songs farthest below their peak rank. Include the song title, artist, current rank, peak rank, `positions_below_peak`, and weeks on chart. Can you identify any patterns? Choose one song from your results and give one plausible explanation for why it may be charting below its earlier peak. Your explanation is a hypothesis, not a conclusion supported by these data alone.

::: source-control
Preview, commit, and push your changes to GitHub with the commit message "Added answer for Question 3".

Make sure to commit and push all changed files so that your source control pane is empty afterward.
:::

## Question 4

Create a scatterplot of `positions_below_peak` (y-axis) versus `weeks_on_chart` (x-axis), and describe the relationship between these two variables. Then, in one pipeline, identify the song that has been on the chart for a long time but is still very close to its peak rank. Include the song title, artist, current rank, peak rank, `positions_below_peak`, and weeks on chart in your output. Finally, comment on where on the scatterplot this song can be found.

## Question 5

*Do songs from some genres tend to have higher weekly stream counts than others?*

To explore this question, create side-by-side boxplots of weekly streams (`streams`) by genre (`genre`). To keep the plot readable, include only the genres with at least 10 songs on the chart.

- How do typical weekly stream counts compare across these genres?

- How do the variabilities of weekly stream counts compare across these genres?

- Which genre contains the single most-streamed song among the genres in your plot? In a single pipeline, identify the song by name and its weekly stream count.

::: callout-tip
In addition to answering the specific prompts above, make sure your response answers the initial question, "Do songs from some genres tend to have higher weekly stream counts than others?"
:::

::: source-control
Now is another good time to preview, commit, and push your changes to GitHub with an informative and concise commit message.

Once again, make sure to commit and push all changed files so that your source control pane is empty afterwards.
:::

## Question 6

*Do some genres have a higher percentage of songs with above-typical weekly stream counts?*

Create a segmented bar plot with one bar per genre and the bar filled with colors according to the value of `stream_level` – one color indicating `Above median` and the other color indicating `At or below median` weekly streams. The y-axis of the segmented barplot should range from 0 to 1, indicating proportions. Compare the percentage of songs with above-median weekly stream counts across the genres based on this plot. Make sure to supplement your narrative with rough estimates of these percentages. Include only genres with at least 10 songs on the plot.

::: callout-tip
For this question, you should begin with the data wrangling pipeline below. We will learn more about data wrangling in the coming weeks, so this is a mini-preview. This pipeline creates a new variable called `stream_level` based on whether a song's weekly streams are above the median of all songs in the dataset. The resulting data frame is assigned back to `spotify`, overwriting the existing data frame with a version that includes the new `stream_level` variable.

```{r}
#| label: hint
#| eval: false
spotify <- spotify |>
  mutate(
    stream_level = if_else(
      streams > median(streams),
      "Above median",
      "At or below median"
    )
  )
```

:::


::: source-control
Now is another good time to preview, commit, and push your changes to GitHub with an informative and concise commit message.

And once again, make sure to commit and push all changed files so that your source control pane is empty afterward. We keep repeating this because it's important and because we see students forget to do this. So take a moment to make sure you're following along with the version control instructions.
:::

# AI use disclosure

{{< include _ai-disclosure.qmd >}}

# Wrap-up

::: source-control
Before you wrap up the assignment, make sure that you preview, commit, and push one final time so that the final versions of both your `.qmd` file and the rendered PDF are pushed to GitHub and your source control pane is empty. We will be checking these to make sure you have been practicing how to commit and push changes.
:::

## Submission

{{< include _submission.qmd >}}

## Grading and feedback

You can get immediate feedback on all of the questions with AI.

Submit your final version of `hw-1.pdf` to Gradescope. Be sure to select the pages associated with each question.

A subset of the questions will be carefully graded for correctness and quality of explanation by the course instructional team. You will receive that feedback in about a week.

There are also workflow points for:

- committing at least three times as you work through your homework,
- having your final version of `.qmd` and `.pdf` files in your GitHub repository, and
- overall organization.
