---
title: HW 3
subtitle: Inflation everywhere
description: "Due: Wed, Sep 23, 11:59 pm"
filters:
  - ../remove-fmt-skip.lua
---

# Introduction

TO DO: Add

## Learning objectives

By the end of this homework, you will:

TO DO: Add

## Getting Started

## Guidelines

The guidelines should feel familiar, as they are the same ones from earlier assignments.
They are included below as a reminder.

{{< include _guidelines.qmd >}}

## AI support and feedback

{{< include _ai-support-feedback.qmd >}}

::: callout-note
If you have not yet enabled Duke's ChatGPT Edu access or installed the Codex extension for Positron, follow the instructions in [HW 1](hw-1.qmd).
:::

# Packages

You will use the **tidyverse** package for data wrangling and visualization.

```{r}
#| label: load-packages
#| eval: true
#| message: false
library(tidyverse)
```

You may choose to load other packages as needed, particularly for improving your plots.
If you do so, add the functions to load these packages in the code cell labeled `load-packages`.

# Questions

## Part 1: Inflation around the world

For this part of the analysis you will work with inflation data from various countries in the world over 30 years.

```{r}
country_inflation <- read_csv("data/country-inflation.csv")
```

### Question 1

Let's first get to know the data.

a.  `glimpse()` at the `country_inflation` data frame and answer the following questions based on the output.
    - How many rows does `country_inflation` have and what does each row represent?
    - How many columns does `country_inflation` have and what does each column represent?
b.  Display the names of the countries included in the dataset. How many distinct countries are there?

::: callout-tip
Find distinct countries with `distinct()` and print all countries with `print(n = ___)` added to your pipeline, where the blank is the number of distinct countries.
:::

::: source-control
Preview, commit, and push your changes to GitHub with the commit message "Added answer for Question 1".
Make sure to commit and push all changed files so that your source control pane is empty afterward.
:::

### Question 2

2024 is the last year for which we have data on United States inflation, so we'll focus on that year for this question.

a.  Which countries had the top three highest inflation rates in 2024?
    Your output must be a data frame with two columns, `country` and `2024`, with inflation rates in descending order, and three rows for the top three countries.

b.  Briefly comment on how the inflation rates for these countries compare to the inflation rate for United States in that year.

::: callout-tip
Column names that are numbers are not considered "proper" in R, therefore to select them you’ll need to surround them with backticks (\`).
:::

::: source-control
Preview, commit, and push your changes to GitHub with an informative and concise commit message.
Make sure to commit and push all changed files so that your source control pane is empty afterward.
:::

### Question 3

Let's next focus on years 1993 and 2025, which are the first and last years in the dataset.

a.  In a single pipeline,

    - filter the `country_inflation` data frame to include only the countries that have inflation data for both 1993 and 2025,
    - then, calculate the ratio of the inflation in 2025 and inflation in 1993 for each country and store this information in a new column called `inf_ratio`,
    - then, select the variables `country` and `inf_ratio`, and
    - store the result in a new data frame called `country_inflation_ratios`.

b.  Then, in two separate pipelines,

    - arrange `country_inflation_ratios` in increasing order of `inf_ratio` and
    - arrange `country_inflation_ratios` in decreasing order of `inf_ratio`.

c.  Which country’s inflation increase is the largest over this time period and by how much?

d.  Which country’s inflation decrease is the largest over this time period and by how much?

::: source-control
Preview, commit, and push your changes to GitHub with an informative and concise commit message.
Make sure to commit and push all changed files so that your source control pane is empty afterward.
:::

### Question 4

Reshape (pivot) `country_inflation` such that each row represents a country/year combination.
Then, display the resulting data frame and write a sentence describing it, including its dimensions (number of rows and columns) and the names of the columns.

::: callout-warning
## Requirements

- Your code must use a pivoting function. There are other ways you can do this reshaping move in R, but this question requires solving this problem by pivoting.
- Your pivoting function must also ensure that in the resulting data frame the year variable is numeric.
- The resulting data frame must be saved as something other than `country_inflation` so you (1) can refer to this data frame later in your analysis and (2) do not overwrite `country_inflation`. You should choose a short but informative name.
:::

::: openai
This would be a good time to get feedback from AI if you haven't yet done so!
Just launch Codex in Positron and ask it to review your answer for Question 4, or any earlier question(s).
:::

### Question 5

Use the pivoted version of the `country_inflation` dataset that you created in Question 4 to answer this questions.

::: callout-warning
## Requirements

- For each part, use a separate, single pipeline.
- Your code must use the `filter()` function for each part, not `arrange()`.
:::

a.  What is the highest inflation rate observed between 1993 and 2025?
    The output of the pipeline must be a data frame with one row and three columns.
    In addition to code and output, your response must include a single sentence stating the country and year.

b.  What is the lowest inflation rate observed between 1993 and 2025?
    The output of the pipeline must be a data frame with one row and three columns.
    In addition to code and output, your response must include a single sentence stating the country and year.

c.  Putting (a) and (b) together: What are the highest and the lowest inflation rates observed between 1993 and 2025?
    The output of the pipeline must be a data frame with two rows and three columns, with the first row corresponding to the highest inflation rate and the second row corresponding to the lowest inflation rate you identified in parts (a) and (b).

::: source-control
Preview, commit, and push your changes to GitHub with an informative and concise commit message.
:::

### Question 6

Once again, use the pivoted version of the `country_inflation` dataset that you created in Question 4 to answer this questions.

a.  Create a vector called `countries_of_interest` which contains the names of up to five countries you want to visualize the inflation rates for over the years. For example, if these countries are Türkiye and United States, you can express this as follows:

```{r}
countries_of_interest <- c("Türkiye", "United States")
```

If they are Türkiye, United States, and Chile, you can express this as follows:

```{r}
# fmt: skip
countries_of_interest <- c("Türkiye", "United States", "Chile")
```

So on and so forth...
Then, in 1-2 sentences, state why you chose these countries.

::: callout-note
Your `countries_of_interest` must consist of no more than five countries.
Make sure that the spelling of your countries matches how they appear in the dataset.
:::

b.  In a single pipeline, filter your pivoted dataset to include only the `countries_of_interest` from part (a), and save the resulting data frame with a new name so you (1) can refer to this data frame later in your analysis and (2) do not overwrite the data frame you're starting with. Use a short but informative name. Then, in a new pipeline, find the `distinct()` countries in the data frame you created.

::: callout-tip
The number of distinct countries in the filtered data frame you created in part (b) must equal the number of countries you chose in part (a).
If it doesn't, you might have misspelled a country name or made a mistake in filtering for these countries.
Go back and correct your work.
:::

c.  Using your data frame from part (b), create a plot of annual inflation vs. year for these countries. Then, in 1-2 sentences, state how you customized your plot.

::: callout-warning
## Requirements

In your plot:

- Data must be represented with points as well as lines connecting the points for each country.
- Each country must be represented by a different color line and different color and shape points.
- Axes and legend must be properly labeled.
- The plot must have an appropriate title (and optionally a subtitle).
- Plot must be customized in at least one way – you could use a different than default color scale, or different than default theme, or some other customization(s).
:::

d.  In a few sentences, describe the patterns you observe in your plot from part (c), particularly focusing on anything you find surprising or not surprising, based on your knowledge (or lack thereof) of these countries' economies.

::: source-control
You know what to do: Preview, commit, and push your changes to GitHub with an informative and concise commit message.
Make sure to commit and push all changed files so that your source control pane is empty afterward.
:::

## Part 2: Inflation in the United States

The OECD defines inflation as follows:

> Inflation is a rise in the general level of prices of goods and services that households acquire for the purpose of consumption in an economy over a period of time.
>
> The main measure of inflation is the annual inflation rate which is the movement of the Consumer Price Index (CPI) from one month/period to the same month/period of the previous year expressed as percentage over time.
>
> Source: [OECD CPI FAQ](https://www.oecd.org/sdd/prices-ppp/consumerpriceindices-frequentlyaskedquestionsfaqs.htm#1)

CPI is broken down into 12 expenditures such as food, housing, health, etc.
Your goal in this part is to create another time series plot of annual inflation, this time for US only.

The data you will need to create this visualization is spread across two files:

- `us-inflation.csv`: Annual inflation rate for the US for 12 CPI expenditures. Each expenditure is identified by an ID number.
- `cpi-expenditures.csv`: A "lookup table" of CPI expenditure ID numbers and their descriptions.

Let's load both of these files.

```{r}
us_inflation <- read_csv("data/us-inflation.csv")
cpi_expenditures <- read_csv("data/cpi-expenditures.csv")
```

### Question 7

a.  How many columns and how many rows does the `us_inflation` dataset have?
    What are the variables in it?
    Which years do these data span?
    Write a brief (1-2 sentences) narrative summarizing this information.

b.  How many columns and how many rows does the `cpi_expenditures` dataset have?
    What are the variables in it?
    Write a brief (1-2 sentences) narrative summarizing this information.

c.  Create a new dataset by joining the `us_inflation` dataset with the `cpi_expenditures` dataset.

- Determine which type of join is the most appropriate one and use that.

- Note that the two datasets don't have a variable with a common name, though they do have variables that contain common information but are named differently.
  You will need to first figure out which variables those are, and then define the `by` argument and use the `join_by()` function to indicate these variables to join the datasets by.

- Use a short but informative name for the joined dataset, and do not overwrite either of the datasets that go into creating it.

Then, find the number of rows and columns of the resulting dataset and report the names of its columns.
Add a brief (1-2 sentences) narrative summarizing this information.

::: source-control
You know what to do!
:::

### Question 8

a.  Create a vector called `expenditures_of_interest` which contains the descriptions or IDs of CPI expenditures you want to visualize. Your `expenditures_of_interest` must consist of no more than five expenditures. If you're using descriptions, make sure that the spelling of your expenditures matches how they appear in the dataset. Then, in 1-2 sentences, state why you chose these expenditures.

::: callout-tip
Refer back to the guidance provided in [Question 6] if you're not sure how to create this vector.
:::

b.  In a single pipeline, filter your joined dataset to include only the `expenditures_of_interest` from part (a), and save the resulting data frame with a new name so you (1) can refer to this data frame later in your analysis and (2) do not overwrite the data frame you're starting with. Use a short but informative name. Then, in a new pipeline, find the `distinct()` expenditures in the data frame you created.

::: source-control
You know what to do!
:::

### Question 9

Using your data frame from the previous question, create a plot of annual inflation vs. year for these expenditures.
Then, in a few sentences, describe the patterns you observe in the plot, particularly focusing on anything you find surprising or not surprising, based on your knowledge (or lack thereof) of inflation rates in the US over the last decade.

::: callout-warning
## Requirements

In your plot:

- Data must be represented with points as well as lines connecting the points for each expenditure.
- Each expenditure must be represented by a different color line and different color and shape points.
- Axes and legend must be properly labeled.
- The plot must have an appropriate title (and optionally a subtitle).
- Plot must be customized in at least one way – you could use a different than default color scale, or different than default theme, or some other customization.
- If your legend has labels that are too long, you can try moving the legend to the bottom and stack the labels vertically. *Hint:* The `legend.position` and `legend.direction` arguments of the `theme()` functions will be useful.

```{r}
ggplot(...) +
  ... +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical"
  )
```
:::

::: openai
This would be a good time to get feedback from AI if you haven't yet done so!
Just launch Codex in Positron and ask it to review your answer for Question 9, or any earlier question(s).
:::

::: source-control
You know what to do!
:::

# AI use disclosure

{{< include _ai-disclosure.qmd >}}

# Wrap-up

::: source-control
Before you wrap up the assignment, make sure that you preview, commit, and push one final time so that the final versions of both your `.qmd` file and the rendered PDF are pushed to GitHub and your source control pane is empty.
We will be checking these to make sure you have been practicing how to commit and push changes.
:::

## Submission

{{< include _submission.qmd >}}

## Grading and feedback

You can get immediate feedback on all of the questions with AI.

Submit your final version of `hw-3.pdf` to Gradescope.
Be sure to select the pages associated with each question.

A subset of the questions will be carefully graded for correctness and quality of explanation by the course instructional team.
You will receive that feedback in about a week.

There are also workflow points for:

- committing at least three times as you work through your homework,
- having your final version of `.qmd` and `.pdf` files in your GitHub repository, and
- overall organization.