---
title: "AE 07: 500 recent opinion articles in The Chronicle"
subtitle: "Suggested answers"
categories: 
  - Application exercise
  - Answers
---

::: callout-important
These are suggested answers.
This document should be used as reference only, it's not designed to be an exhaustive key.
:::

## Part 1 - Data scraping

Fill in `ae-07-chronicle-scrape.R`for suggested scraping code.

## Part 2 - Data analysis

Let's start by loading the packages we will need:

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
```

-   Load the data you saved into the `data` folder and name it `chronicle`.

```{r}
#| label: load-data
#| message: false
chronicle <- read_csv("data/chronicle-2026-09-29.csv")
```

-   How many opinion articles were published in The Chronicle since the beginning of the semester?

```{r}
#| label: count-semester-articles
chronicle |>
  filter(date >= ymd("2026-08-24")) |>
  nrow()
```

-   Who are the top 10 most prolific authors among the 500 most recent opinion articles and how many articles have they each published?

```{r}
#| label: count-authors-articles
chronicle |>
  count(author, sort = TRUE)
```

-   Display the date, title, and author of any article(s) that mention AI in their title.

```{r}
#| label: ai-title
chronicle |>
  filter(str_detect(title, "AI")) |>
  select(date, title, author)
```

-   Come up with another question and try to answer it using the data.

```{r}
#| label: add-label
# add code here
```
