---
title: "AE 03: Tidying Stat Sci majors"
categories: 
  - Application exercise
  - Answers
---

::: callout-important
These are suggested answers.
This document should be used as reference only, it's not designed to be an exhaustive key.
:::

## Data

The data come from the [Office of the University Registrar](https://registrar.duke.edu/registration/enrollment-statistics).

```{r}
#| label: load-packages-data
#| message: false

library(tidyverse)

majors <- read_csv("data/majors.csv")

statsci_majors <- majors |>
  filter(
    department == "Statistical Science",
    academic_plan != "Minor"
  )
```

And let's take a look at the data.

```{r}
statsci_majors
```

## Pivoting

-   **Demo:** Pivot the `statsci` data frame *longer* such that:

    -   Each row represents a degree type / year combination
    -   `year` and `n`umber of graduates for that year are columns in the data frame
    -   The resulting `year` column is numeric

```{r}
#| label: pivot
statsci_majors |>
  pivot_longer(
    cols = !c(department, academic_plan),
    names_to = "year",
    values_to = "n",
    names_transform = as.numeric
  )
```

-   **Your Turn:** Now, repeat your code from above, but this time **save the result** to a **new** variable name.

```{r}
#| label: pivot-with-transform-name
statsci_majors_longer <- statsci_majors |>
  pivot_longer(
    cols = !c(department, academic_plan),
    names_to = "year",
    names_transform = as.numeric,
    values_to = "n"
  )
```

## Plotting

-   **Your turn:** Now we will start making our plot, but let's not get too fancy right away. Just get the geoms right for now and use default colors for different academic plans.

```{r}
#| label: plot-draft
statsci_majors_longer |>
  ggplot(aes(x = year, y = n, color = academic_plan)) +
  geom_point() +
  geom_line()
```

-   **Question**: Why was the pivot necessary in order to create this plot?

*Add your response here!*

-   **Question:** What aspects of the plot need to be updated still?

*Add your response here.*

-  **Demo:** Update `statsci_majors_longer` to fix the discontinuities.

```{r}
#| label: pivot-again
statsci_majors_longer <- statsci_majors_longer |>
  mutate(n = if_else(is.na(n), 0, n))
```

-   **Demo:** Update x-axis scale such that the years displayed go from 2011 to 2025 in increments of 2 years. Do this by adding on to your pipeline from earlier.

```{r}
#| label: plot-improve-1
statsci_majors_longer |>
  ggplot(aes(x = year, y = n, color = academic_plan)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = seq(2011, 2025, 2))
```

-   **Demo:** Update line colors using the following level / color assignments. Once again, do this by adding on to your pipeline from earlier.
    -   "BS" = "cadetblue4"

    -   "BS2" = "cadetblue3"

    -   "BA" = "lightgoldenrod4"

    -   "BA2" = "lightgoldenrod3"

```{r}
#| label: plot-improve-2
statsci_majors_longer |>
  ggplot(aes(x = year, y = n, color = academic_plan)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = seq(2011, 2025, 2)) +
  scale_color_manual(
    values = c(
      "BS" = "cadetblue4",
      "BS2" = "cadetblue3",
      "BA" = "lightgoldenrod4",
      "BA2" = "lightgoldenrod3"
    )
  )
```

-   **Your turn:** Update the plot labels (`title`, `subtitle`, `x`, `y`, and `caption`) and use `theme_minimal()`. Once again, do this by adding on to your pipeline from earlier.

```{r}
#| label: plot-improve-3
statsci_majors_longer |>
  ggplot(aes(x = year, y = n, color = academic_plan)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = seq(2011, 2025, 2)) +
  scale_color_manual(
    values = c(
      "BS" = "cadetblue4",
      "BS2" = "cadetblue3",
      "BA" = "lightgoldenrod4",
      "BA2" = "lightgoldenrod3"
    )
  ) +
  labs(
    x = "Graduation year",
    y = "Number of majors graduating",
    color = "Degree type",
    title = "Statistical Science majors over the years",
    subtitle = "Academic years 2012 - 2026",
    caption = "Source: Office of the University Registrar\nhttps://registrar.duke.edu/registration/enrollment-statistics"
  ) +
  theme_minimal()
```

-   **Demo:** Move the legend into the plot, make its background white, and its border gray.

```{r}
#| label: plot-improve-4
statsci_majors_longer |>
  ggplot(aes(x = year, y = n, color = academic_plan)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = seq(2011, 2025, 2)) +
  scale_color_manual(
    values = c(
      "BS" = "cadetblue4",
      "BS2" = "cadetblue3",
      "BA" = "lightgoldenrod4",
      "BA2" = "lightgoldenrod3"
    )
  ) +
  labs(
    x = "Graduation year",
    y = "Number of majors graduating",
    color = "Degree type",
    title = "Statistical Science majors over the years",
    subtitle = "Academic years 2012 - 2026",
    caption = "Source: Office of the University Registrar\nhttps://registrar.duke.edu/registration/enrollment-statistics"
  ) +
  theme_minimal() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.1, 0.7),
    legend.background = element_rect(fill = "white", color = "grey")
  )
```

-   **Demo:** Finally, set `fig-width: 8` and `fig-height: 5` for your plot in the chunk options.

```{r}
#| label: plot-improve-5
#| fig-width: 8
#| fig-height: 5
statsci_majors_longer |>
  ggplot(aes(x = year, y = n, color = academic_plan)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = seq(2011, 2025, 2)) +
  scale_color_manual(
    values = c(
      "BS" = "cadetblue4",
      "BS2" = "cadetblue3",
      "BA" = "lightgoldenrod4",
      "BA2" = "lightgoldenrod3"
    )
  ) +
  labs(
    x = "Graduation year",
    y = "Number of majors graduating",
    color = "Degree type",
    title = "Statistical Science majors over the years",
    subtitle = "Academic years 2012 - 2026",
    caption = "Source: Office of the University Registrar\nhttps://registrar.duke.edu/registration/enrollment-statistics"
  ) +
  theme_minimal() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.1, 0.7),
    legend.background = element_rect(fill = "white", color = "grey")
  )
```
