---
title: "AE 04: Sales taxes + data joining"
subtitle: Suggested answers
categories: 
  - Application exercise
  - Answers
---

::: callout-important
These are suggested answers.
This document should be used as a reference only; it's not designed to be an exhaustive key.
:::

# Getting started

## Packages

We'll use the **tidyverse** package for this analysis.

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
```

## Data

The data are available in the `data` folder.

```{r}
#| label: load-data
#| include: false
sales_taxes <- read_csv(here::here("ae", "data/sales-taxes-26.csv"))
us_regions <- read_csv(here::here("ae", "data/us-regions.csv"))
```

```{r}
#| label: load-data-show
#| eval: false
sales_taxes <- read_csv("data/sales-taxes-26.csv")
us_regions <- read_csv("data/us-regions.csv")
```

And let's take a look at the data.

```{r}
#| label: glimpse
glimpse(sales_taxes)
glimpse(us_regions)
```

# Joining

Join the sales tax data with region data and save the joined data frame as a new data frame, not overwriting either data frame that goes into the join.

```{r}
#| label: join
sales_taxes_regions <- left_join(
  sales_taxes,
  us_regions,
  by = join_by(state == state_name)
)
```

Calculate the average sales tax of states in each region.
What is surprising in the output?

```{r}
#| label: region-means-take-1
sales_taxes_regions |>
  group_by(region) |>
  summarize(mean_state_tax = mean(state_tax_rate))
```

Identify the state with `NA` for region.

```{r}
#| label: identify-na
sales_taxes_regions |>
  filter(is.na(region)) |>
  select(state, region)
```

Suprised? 
You shouldn't be...
Read [here](https://en.wikipedia.org/wiki/No_taxation_without_representation#District_of_Columbia) for more information on why you shouldn't be surprised.

Apply a fix for the `NA` in region, and calculate the mean sales taxes for regions again.
Display the results in ascending order of mean sales tax.

```{r}
#| label: region-means-take-2
sales_taxes_regions <- sales_taxes_regions |>
  mutate(
    region = if_else(state == "District of Columbia", "Northeast", region)
  )

sales_taxes_regions |>
  group_by(region) |>
  summarize(mean_state_tax = mean(state_tax_rate)) |>
  arrange(mean_state_tax)
```

# Joining another way

Accomplish the same task using a different join this time.

```{r}
#| label: another-join
sales_taxes_regions_2 <- right_join(
  us_regions,
  sales_taxes,
  by = join_by(state_name == state)
)

sales_taxes_regions_2 <- sales_taxes_regions_2 |>
  mutate(
    region = if_else(state_name == "District of Columbia", "Northeast", region)
  )

sales_taxes_regions_2 |>
  group_by(region) |>
  summarize(mean_state_tax = mean(state_tax_rate)) |>
  arrange(mean_state_tax)
```
