---
title: "AE 07: 500 recent opinion articles in The Chronicle"
categories: 
  - Application exercise
---

## Part 1 - Data scraping

Fill in `ae-07-chronicle-scrape.R`for suggested scraping code.

## Part 2 - Data analysis

Let's start by loading the packages we will need:

```{r}
#| label: load-packages
# add code here
```

-   Load the data you saved into the `data` folder and name it `chronicle`.

```{r}
#| label: load-data
# add code here
```

-   How many opinion articles were published in The Chronicle since the beginning of the semester?

```{r}
#| label: count-semester-articles
# add code here
```

-   Who are the top 10 most prolific authors among the 500 most recent opinion articles and how many articles have they each published?

```{r}
#| label: count-authors-articles
# add code here
```

-   Display the date, title, and author of any article(s) that mention AI in their title.

```{r}
#| label: ai-title
# add code here
```

-   Come up with another question and try to answer it using the data.

```{r}
#| label: add-label
# add code here
```
