---
title: "AE 03: Tidying Stat Sci majors"
subtitle: "Suggested answers"
categories: 
  - Application exercise
  - Answers
---

::: callout-important
These are suggested answers.
This document should be used as reference only, it's not designed to be an exhaustive key.
:::

## Data

The data come from the [Office of the University Registrar](https://registrar.duke.edu/registration/enrollment-statistics).

```{r}
#| label: load-packages-data
#| message: false

library(tidyverse)

majors <- read_csv("data/majors.csv")

statsci_majors <- majors |>
  filter(
    department == "Statistical Science",
    academic_plan != "Minor"
  )
```

And let's take a look at the data.

```{r}
statsci_majors
```

## Pivoting

-   **Demo:** Pivot the `statsci` data frame *longer* such that:

    -   Each row represents a degree type / year combination
    -   `year` and `n`umber of graduates for that year are columns in the data frame
    -   The resulting `year` column is numeric

```{r}
#| label: pivot
# add code here
```

-   **Your Turn:** Now, repeat your code from above, but this time **save the result** to a **new** variable name.

```{r}
#| label: pivot-with-transform-name
# add code here
```

## Plotting

-   **Your turn:** Now we will start making our plot, but let's not get too fancy right away. Just get the geoms right for now and use default colors for different academic plans.

```{r}
#| label: plot-draft
# add code here
```

-   **Question**: Why was the pivot necessary in order to create this plot?

*Add your response here!*

-   **Question:** What aspects of the plot need to be updated still?

*Add your response here.*

-  **Demo:** Update `statsci_majors_longer` to fix the discontinuities.

```{r}
#| label: pivot-again
# add code here
```

-   **Demo:** Update x-axis scale such that the years displayed go from 2012 to 2026 in increments of 2 years. Do this by adding on to your pipeline from earlier.

```{r}
#| label: plot-improve-1
# add code here
```

-   **Demo:** Update line colors using the following level / color assignments. Once again, do this by adding on to your pipeline from earlier.
    -   "BS" = "cadetblue4"

    -   "BS2" = "cadetblue3"

    -   "BA" = "lightgoldenrod4"

    -   "BA2" = "lightgoldenrod3"

```{r}
#| label: plot-improve-2
# add code here
```

-   **Your turn:** Update the plot labels (`title`, `subtitle`, `x`, `y`, and `caption`) and use `theme_minimal()`. Once again, do this by adding on to your pipeline from earlier.

```{r}
#| label: plot-improve-3
# add code here
```

-   **Demo:** Move the legend into the plot, make its background white, and its border gray.

```{r}
#| label: plot-improve-4
# add code here
```

-   **Demo:** Finally, set `fig-width: 8` and `fig-height: 5` for your plot in the chunk options.

```{r}
#| label: plot-improve-5
#| fig-width: 8
#| fig-height: 5
# add code here
```
