---
title: "AE 04: Sales taxes + data joining"
categories: Application exercise
---

# Getting started

## Packages

We'll use the **tidyverse** package for this analysis.

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
```

## Data

The data are available in the `data` folder.

```{r}
#| label: load-data-show
#| message: false
sales_taxes <- read_csv("data/sales-taxes-26.csv")
us_regions <- read_csv("data/us-regions.csv")
```

And let's take a look at the data.

```{r}
#| label: glimpse
# add code here
```

# Joining

Join the sales tax data with region data and save the joined data frame as a new data frame, not overwriting either data frame that goes into the join.

```{r}
#| label: join
# add code here
```

Calculate the average sales tax of states in each region.
What is surprising in the output?

```{r}
#| label: region-means-take-1
# add code here
```

Identify the state with `NA` for region.

```{r}
#| label: identify-na
# add code here
```

Suprised? 
You shouldn't be...
Read [here](https://en.wikipedia.org/wiki/No_taxation_without_representation#District_of_Columbia) for more information on why you shouldn't be surprised.

Apply a fix for the `NA` in region, and calculate the mean sales taxes for regions again.
Display the results in ascending order of mean sales tax.

```{r}
#| label: region-means-take-2
# add code here
```

# Joining another way

Accomplish the same task using a different join this time.

```{r}
#| label: another-join
# add code here
```
