---
title: Lab 3
subtitle: To be posted
description: "Due: End of lab on Thurs, Sep 17"
categories: "Lab" 
draft: true
execute:
  eval: false
---

::: {.callout-important}
## Find your teammates
... and sit with them!
:::

# Introduction

This lab is a deep dive into `group_by()`!

Make sure to upload your completed lab to Gradescope **by the end of your lab session** and commit and push your final version to GitHub.

## Learning objectives

By the end of this lab, you will:

- TO DO: Add

# Getting started

Clone your **lab-2** repository using the same process used in Lab 0. 

1. Go to the course GitHub organization at [github.com/sta199-f26](https://github.com/sta199-f26) and click on the repository with the prefix `lab-3`. 
2. Select the green **Code** button and choose **HTTPS** under **Clone**. Click on the clipboard icon to copy the repo URL.
3. In Positron's Welcome window, click on *New Folder From Git...*.
4. Paste the repository URL into the dialog box labeled *Git repository URL* and select **OK**.
5. Open `lab-2.qmd` in the Explorer pane.

Then, click *lab-3.qmd* to open the template Quarto file and update the `authors` field to add your name first (first and last) and then your teammates' names (first and last).
Render the document.
If you get a popup window error, click "Try again".
Examine the rendered document and make sure your name is updated in the document.
Commit and push your changes with a meaningful commit message and push to GitHub.

## Guidelines

{{< include _guidelines.qmd >}}

# Packages

In this lab we will work with the **tidyverse** package.

```{r}
#| label: load-packages
#| eval: true
#| message: false
library(tidyverse)
```

-   **Run** the code cell by clicking on the green triangle (play) button for the code cell labeled `load-packages`. This loads the package to make its features (the functions and datasets in it) accessible from your *Console*.
-   Then, **preview** the document, which loads this package to make its features (the functions and datasets in it) available for other code cells in your Quarto document.

# Questions

## Question 1

For this question, you will work with the following dataset called `scores`.

```{r}
#| eval: true
scores <- tribble(
  ~student_id , ~math , ~english ,
  "S1"        ,    90 ,       92 ,
  "S2"        ,    85 ,       80 ,
  "S3"        ,    88 ,       85 ,
  "S4"        ,    95 ,       74
)

scores
```

::: callout-note
The `tribble()` function is helpful for creating small data frames (`tibble`s) with an easier to read row-by-row layout.
:::

It has three variables (`student_id`, `math`, and `english`) and four rows (one for each student).

Before writing any code, answer the following questions:

- Suppose you want to reshape the data frame so that there is one row per student per subject.
  What function would you use to do this?

- If you reshaped the data to have one row per student per subject, 
  - how many rows would the resulting data frame have?
  - how many columns would the resulting data frame have and what would the column names be?

Then, write the code to reshape the data frame as described above. 

::: source-control
Preview, commit, and push your changes to GitHub with a succinct and informative commit message.

Make sure to commit and push all changed files so that your source control pane is empty afterward.
:::

## Question 2

For this question, you will work with the following dataset called `patients`.

```{r}
#| eval: true
patients <- tribble(
  ~patient_id , ~measurement_time_ , ~systolic_bp ,
  "P1"        , "Morning"          ,          120 ,
  "P1"        , "Noon"             ,          115 ,
  "P1"        , "Evening"          ,          123 ,
  "P2"        , "Morning"          ,          118 ,
  "P2"        , "Evening"          ,          121
)

patients
```

It has three variables (`patient_id`, `measurement_time_`, and `systolic_bp` -- short for systolic blood pressure) and five rows (one per patient per measurement time).

Before writing any code, answer the following questions:

- Suppose you want to reshape the data frame so that there is one row per patient and measurements at different times of the day are recorded in different columns.
  What function would you use to do this?

- If you reshaped the data to have one row per patient, 
  - how many rows would the resulting data frame have?
  - how many columns would the resulting data frame have and what would the column names be?

Then, write the code to reshape the data frame as described above. What does the `NA` value mean in the resulting data frame?

::: source-control
Preview, commit, and push your changes to GitHub with a meaningful commit message.

Make sure to commit and push all changed files so that your source control pane is empty afterward.
:::

## Question 3

For this question, you will work with the following dataset called `quiz_scores`.

```{r}
#| eval: true
quiz_scores <- tribble(
  ~student_id , ~week_1 , ~week_2 , ~week_3 ,
  "S1"        , "80"    , "85"    , "90"    ,
  "S2"        , "75"    , NA      , "88"    ,
  "S3"        , "92"    , "90"    , "95"
)

quiz_scores
```

It has one row per student and one column for each week's quiz score.
Notice that the scores are stored as character strings (in quotation marks), and one score is missing.

You want to reshape this data frame to have one row per student per week and three columns: `student_id`, `week`, and `score`.
The `week` column should contain the integers `1`, `2`, and `3`, and `score` should be numeric.
Keep the row corresponding to the missing score.

a. Before writing any code, answer the following questions:

  - What function would you use to reshape the data, and how many rows would the resulting data frame have?
  - If you only specified which columns to pivot and the new column names, what values and data type would you expect in `week`? What data type would you expect in `score`?

Then, complete the following steps:

b. Read about `names_prefix`, `names_transform`, and `values_transform` in the documentation for `pivot_longer()`. Reshape `quiz_scores` as described above using these arguments within a single `pivot_longer()` call, and save the result as `quiz_scores_long`.

::: callout-tip
Hint: `as.integer()` converts values to integers, and `as.numeric()` converts values to numbers. Supply transformation functions without parentheses, using the form `list(column_name = function_name)`.
:::

c. Display `quiz_scores_long` and use `glimpse()` to check the column types.Explain what each of the three new arguments does and why removing the prefix alone is not enough to make `week` an integer.

d. Reshape `quiz_scores_long` back to one row per student, with columns named `week_1`, `week_2`, and `week_3`. Use `names_prefix` in `pivot_wider()` to restore the prefix. How does the role of `names_prefix` differ between the two pivot functions? How do the score column types in this result compare with those in the original `quiz_scores`?

::: source-control
Preview, commit, and push your changes to GitHub with a meaningful commit message.

Make sure to commit and push all changed files so that your source control pane is empty afterward.
:::

# Wrap-up

## Submission

{{< include _submission.qmd >}}

## Grading and feedback

This lab is worth 30 points:

- 10 points for attending lab and submitting work by the end of lab; and
- 20 points for correct and complete answers, clear presentation, and following the required workflow.

The workflow portion includes making at least three meaningful commits, pushing the final `.qmd` and rendered PDF to GitHub, and keeping your work organized.
